---
title: "hydrochem_clustering_Excel"
author:  
date:
geometry: margin=2cm
documentclass: "report"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
```

```{r Install packages, eval=FALSE, include=FALSE}
#Install packages ( only needed when using the script for the first time!)
install.packages(c("readxl","PeriodicTable","knitr","ggplot2", "ggmap","plotly","factoextra","tibble","cluster","purr","tidyr", "magrittr", "scales","reshape2","directlabels", "dplyr", "hydrogeo", "ggrepel","readxl","ggplot2", "knitr","PeriodicTable","RMySQL","devtools", "latex2exp","devtools", "ggforce","kableExtra"))

devtools::install_github("kassambara/ggpubr")
```

```{r Working directory, include=FALSE}
# By default the working directory is set below to be the folder containing the RMarkdown file
wd = getwd()
setwd(wd)
```

```{r Provide the instructions for your run, message=FALSE, warning=FALSE, include=FALSE}

#▐▓█▀▀▀▀▀▀▀▀▀█▓▌░▄▄▄▄▄░
#▐▓█░░▀░░▀▄░░█▓▌░█▄▄▄█░
#▐▓█░░▄░░▄▀░░█▓▌░█▄▄▄█░
#▐▓█▄▄▄▄▄▄▄▄▄█▓▌░█████░
#░░░░▄▄███▄▄░░░░░█████░


#1) Provide the path to the .xlsx file containing your analyses  _________________________________________#                                 
library(readxl)
water_analysis <-
    read_excel("Example.xlsx")

#2)Clustering

#Choose the similarity measure to be used 
#( "manhattan", "euclidean", "maximum", "canberra", "binary", "minkowski")
sim_meas = "manhattan" 
# Choose the agglomeration method 
#("complete", "average", "single", "ward.D")
agg_method =  "ward.D" 
# Choose the number of clusters for the dendrogram 
numberofclusters = 4

#3)Specify map corners
map_cor = c(
  top =    37.789129, # Latitude of the upper left corner
  left =   26.798606, # Longitude of the upper left corner
  bottom =   37.655869, # Latitude of the lower right corner
  right = 27.037566  # Longitude of the lower right corner
  )

#___________________________________________YOU CAN NOW KNIT THE DOCUMENT_________# 
```

# 1.Data quality assessment

## 1.1. Data overview

```{r View your original data, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
kable(water_analysis[1:11], align = 'c', caption = 'The studied water analyses ($mg/L$)')
```

## 1.2. Charge balance error

An assessment of the hydrogeochemical data at hand is crucial for the quality of the following modeling procedure and therefore constitutes the first processing stage. A routine validation is checking the **electroneutrality** of the water. Aqueous solutions must be electrically neutral hence the total sum of all the positive charges (cations) must equal the total sum of all negative charges (anions):

$$ \sum eq_{c a t i o n s}=\sum eq_{a n i o n s}$$

$$
\sum m_c * z_c =\ \sum m_a*|z_a|
$$

Where $eq_{ions}$ is in $meq/L$, therefore the original data was transformed from accordingly from $mg/L$ to $meq/L$.

```{r Transform your data from mg/l to meq/l, echo=TRUE, message=FALSE, warning=FALSE}
library("PeriodicTable")

#Basic ions
Na_meq = c(water_analysis$Na / mass('Na') * 1)
K_meq = c(water_analysis$K / mass('K') * 1)
Mg_meq = c(water_analysis$Mg / mass('Mg') * 2)
Ca_meq = c(water_analysis$Ca / mass('Ca') * 2)
Cl_meq = c(water_analysis$Cl / mass('Cl') * 1)
SO4_meq = c(water_analysis$SO4 / (mass("S") + 4 * mass("O")) * 2)
HCO3_meq = c(water_analysis$HCO3 / (mass("H") + mass("C") + 3 * mass("O")) *1)

# Further ions 
if("NO3" %in% colnames(water_analysis)) {
  NO3_meq = c(water_analysis$NO3 / (mass("N") + 3 * mass("O")) * 1)
}

if("Br" %in% colnames(water_analysis)) {
  Br_meq = c(water_analysis$Br / (mass("Br")) * 1)
}

if("CO3" %in% colnames(water_analysis)) {
  CO3_meq = c(water_analysis$CO3 / (mass("C") + 3 * mass("O")) * 2)
}
```

The use of the **Charge Balance Error (CBE)** quantifies the balance of the analysis. Positive CBE values indicate more anions than cations and vice versa. $$CBE = \frac{\sum anions -\sum cations}{\sum anions+\sum cations} * 100$$

Charge Balance Error was calculated for every sample as presented in Table 2.

```{r Charge Balance Error calculation, echo=TRUE, message=FALSE, warning=FALSE}


# Provide the anions and cations to be incooroporated in the Charge Balance Error calculation
cations = Na_meq + K_meq + Mg_meq + Ca_meq
anions = Cl_meq + SO4_meq + HCO3_meq

if(exists("NO3_meq")){
  anions = anions + NO3_meq
}

if(exists("Br_meq")){
  anions = anions + Br_meq
}

if(exists("CO3_meq")){
  anions = anions + CO3_meq
}

# Calculate the charge balance error
C_B_E = (cations - anions) / (cations + anions) * 100

# Add a column with the CBE to the dataset
analysis_CBE <- water_analysis
analysis_CBE$CBE = C_B_E

# Please make sure that the sample name is the first column
library(knitr)
kable(analysis_CBE[, c(1, ncol(analysis_CBE))], align = 'c', 
      caption = 'Water samples and their respective CBE(%)')
```

Systematic errors in the sampling procedure and the overall number of suitable or not samples, can be evaluated by plotting the distribution of the CBE error.

```{r Distribution of the CBE error,fig.cap="Distribution of CBE", echo=TRUE, fig.height=2, fig.width=8, message=FALSE, warning=FALSE}

library(ggplot2)

plot1_density = ggplot(analysis_CBE, aes(CBE)) +
  geom_density(color = "darkblue", fill = "lightblue") +
  geom_vline(
    aes(xintercept = 5),
    size = 1,
    color = "red",
linetype = "dashed"
) +
  geom_vline(
aes(xintercept = -5),
 size = 1,
 color = "red",
linetype = "dashed"
) +
  theme_light() +
  xlab("CBE %") +
  ylab("Density")

plot1_density 
```

Samples with CBE higher than ±5% are presented in Table 3

```{r View samples in which |CBE| exceeds 5%, echo=TRUE}
Large_imbalance=analysis_CBE[abs(analysis_CBE$CBE) > 5,]

library(knitr)
kable(Large_imbalance[1],align = 'c',  caption = 'Samples that exceed the 5% CBE')
```

Samples with a charge imbalance up to ±5%, as indicated by the CBE, are accepted otherwise the sample is dismissed from the dataset and any further analysis.

```{r Exclude samples with high charge imbalance, eval=FALSE, include=FALSE}
water_analysis = analysis_CBE[abs(analysis_CBE$CBE) < 5,]
```

The filtered dataset is transformed from $mg/L$ to $meq/L$

```{r Recalculate meq/L for the filtered dataset, echo=TRUE, message=FALSE, warning=FALSE}
library("PeriodicTable")

### Please notice that this chunk should be identical with the {r Transform your data from mg/l to meq/l} 

#Basic ions
Na_meq = c(water_analysis$Na / mass('Na') * 1)
K_meq = c(water_analysis$K / mass('K') * 1)
Mg_meq = c(water_analysis$Mg / mass('Mg') * 2)
Ca_meq = c(water_analysis$Ca / mass('Ca') * 2)
Cl_meq = c(water_analysis$Cl / mass('Cl') * 1)
SO4_meq = c(water_analysis$SO4 / (mass("S") + 4 * mass("O")) * 2)
HCO3_meq = c(water_analysis$HCO3 / (mass("H") + mass("C") + 3 * mass("O")) *1)

# Further ions 

if("NO3" %in% colnames(water_analysis)) {
  NO3_meq = c(water_analysis$NO3 / (mass("N") + 3 * mass("O")) * 1)
}


if("Br" %in% colnames(water_analysis)) {
  Br_meq = c(water_analysis$Br / (mass("Br")) * 1)
}

if("CO3" %in% colnames(water_analysis)) {
  CO3_meq = c(water_analysis$CO3 / (mass("C") + 3 * mass("O")) * 2)
}

```

The amount of TDS ( Total dissolved ions) in the water samples can be computed and added to the dataset.

```{r Theoretical computation of TDS, echo=TRUE}

water_analysis <- transform(water_analysis, TDS = (Na + K + Mg + Ca + Cl + HCO3+ SO4))

if("NO3" %in% colnames(water_analysis)) {
  water_analysis$TDS = water_analysis$TDS + water_analysis$NO3
}

if("Br" %in% colnames(water_analysis)) {
  water_analysis$TDS = water_analysis$TDS + water_analysis$Br
}

if("CO3" %in% colnames(water_analysis)) {
   water_analysis$TDS = water_analysis$TDS + water_analysis$CO3
}
```

```{r First Housekeeping, include=FALSE}

# Housekeeping! 
rm(list=c("anions","cations","analysis_CBE","C_B_E","analysis_CBE","Large_imbalance","plot1_density"))
```

------------------------------------------------------------------------

# 2. Cluster analysis

## 2.1. Data preparation

The variables upon which the clustering will be performed are chosen.

```{r Create data subset to perform cluster analysis, echo=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
library(magrittr)
library(knitr)
library(tidyr)
library(purrr)

# Choose the ions or variables to be used in the cluster analysis 
# from the water_analysis dataframe

water_analysis_clust = water_analysis %>% 
  select(Sample,Cl,HCO3,SO4,Ca,Mg,Na,K)

if("NO3" %in% colnames(water_analysis)) {
  water_analysis_clust$NO3 = water_analysis$NO3
}


if("Br" %in% colnames(water_analysis)) {
  water_analysis_clust$Br = water_analysis$Br
}


if("CO3" %in% colnames(water_analysis)) {
  water_analysis_clust$CO3 = water_analysis$CO3
}
```

The skewness of the samples is examined by plotting the distribution of the ions.

```{r Examine the skewness, echo=TRUE, fig.height=4, fig.width=8, fig.cap= "Distribution of ions prior to normalization"}
library(dplyr)
library(ggplot2)
library(tibble)

### Look at the distribution of the ions
water_analysis_clust %>% 
  keep(is.numeric) %>% 
  gather() %>%
  ggplot(aes(value)) +                     
    facet_wrap(~ key, scales = "free") +   
    geom_density(color="darkblue", fill="lightblue") + theme_light()+ xlab("")
```

Departure from a more "Normal" distribution was mitigated by applying a $log_{10}$ transformation to all ions as shown in the plot below. The relative differences and ratios are not altered whilst the range of numeric values decreases.

```{r log10 transform, echo=TRUE, fig.height=4, fig.width=8, fig.cap= "Distribution of ions after $log_{10}$ transformation"}

library(dplyr)
library(ggplot2)
library(tibble)

# We assign the samples' names column to the data frame's row names so as to maintain them in the following scaling
water_analysis_clust = water_analysis_clust%>% 
  remove_rownames %>% 
  column_to_rownames(var="Sample")

water_analysis_clust <- log10(water_analysis_clust)


water_analysis_clust %>% 
  keep(is.numeric) %>% 
  gather() %>%
  ggplot(aes(value)) +                     
    facet_wrap(~ key, scales = "free") +   
    geom_density(color="darkblue", fill="lightblue") + theme_light()+ xlab("")

```

To ensure a common scale for all variables, the final data preparation step includes the scaling of the samples substituting each value with a respective Z-score. After scaling, the mean is set to 0 and the distance from the mean is in units of 1 standard deviation:

$$
z = \frac{x - \bar{x}}{\sigma}
$$

The resulting z-scores, to be used in the following cluster analysis are presented in the table below:

```{r z-scoring, echo=TRUE, message=FALSE, warning=FALSE}
library(cluster)
library(dplyr)
library(magrittr)
library(knitr)
library(tidyr)
library(purrr)
library(tibble)

# Actual scaling using the scale() function
water_analysis_clust_scaled <- scale(water_analysis_clust)

# Print the scaled data
kable(water_analysis_clust_scaled,align = 'c',  caption = 'Scaled data Z-scores')
```

## 2.2.Clustering

After having scaled the dataset we can now calculate the dissimilarity matrix (the distance $\delta_{i,j}$ between each pair of observations). Here the **Manhattan** or **City Block Distance (CDB)** was used as a measure of similarity . Other similarity measures can be used substituting the `method =` parameter with `“euclidean”, “maximum”, “canberra”, “binary”, “minkowski”` respectively. In this matrix, the value represent the distance between objects. The values on the diagonal of the matrix represent the distance between objects and themselves (which are zero). We can perform agglomerative HC with `hclust` specify the agglomeration method to be used (i.e. "complete", "average", "single", "ward.D"). We can then plot the dendrogram.

Notice that the measure of similarity and agglomeration method are already defined. You may change them again below to cancel the initial instructions.

```{r HCA, echo=TRUE, fig.cap= " Cluster Dendrogram (ward.D agglomeration method)"}

# Measure of similarity (method = " ")
d <- dist(water_analysis_clust_scaled, method = sim_meas)

# Agglomeration method = " " 
hc1 <- hclust(d, method = agg_method )

plot(hc1, cex = 0.6, hang = -1)
```

For determining an optimal number of clusters the results of the Elbow method are presented below:

```{r cluster optimal number, echo=TRUE, fig.height=3, fig.width=8, fig.cap=""}

library(factoextra)

par(mfrow=c(1,3))

fviz_nbclust(water_analysis_clust_scaled, FUN = hcut, method = "wss")
```

## 3.3.1. Dendrogram based on chosen number of clusters

Based on the number of clusters chosen the previous dendrogram is respectively colored to portray the clustering. Again the number of k mean clusters were defined however, you may change them again below to cancel the initial instructions.

```{r cut the dendrogram, echo=TRUE, fig.height=8, fig.cap=" Dendrogram with the defined number of clusters indicated by the respective colors"}

library(factoextra)

# The number of clusters has been assigned in the first section of instructions.  
# If you want to change the number of clusters activate the assignment below by deleting the "#" below
# numberofclusters = 1 

dendro= fviz_dend(hc1, cex = 0.5, k = numberofclusters, horiz = T,main = "Samos dendrogram (Ward's method)",
          xlab = "", ylab = "Distance", color_labels_by_k = T, palette =  "aaas")

ggsave(file="Dendrogram.svg", dendro, width=10, height=8)

dendro
```

Each sample's respective cluster value is added as a column in `water_analysis` dataframe.

```{r add clusters to data, echo=TRUE}

clusterCut <- cutree(hc1,numberofclusters) 

water_analysis= cbind(water_analysis, Cluster = clusterCut )
```

## 3.3.2. Study area map with the respective clusters

The spatial context of the clustering process was visualized:

```{r create map, echo=FALSE, fig.cap=", fig.height=8, fig.width=10, message=FALSE, warning=FALSE}
library(ggmap)
library(ggrepel)

map_data = data.frame(SAMPLES = water_analysis$Sample,
                      LON = water_analysis$Longitude,
                      LAT = water_analysis$Latitude,
                      CLUSTERS = water_analysis$Cluster)
map_data$CLUSTERS = factor(map_data$CLUSTERS)

myMap <- get_stamenmap(bbox = map_cor,
                       maptype = "terrain", 
                       crop = TRUE,
                       zoom = calc_zoom(LON, LAT, map_data))

 map=ggmap(myMap, darken=c(0.05, "white"), extent="device") + 
  geom_point(aes(x=LON, y=LAT, color=CLUSTERS), data=map_data, size= 3, alpha = 0.8)+
  geom_text_repel(data = map_data, aes(x = LON, y = LAT, label = SAMPLES), size = 3)+
  theme(legend.position="bottom")
 
 map
 
 ggsave(file="Cluster_map.png", map)
```

```{r Export water_analysis with clusters for GIS, include=FALSE}

hydroExport = "/HydroWorkflow_export.csv"

gispath = c(wd,hydroExport)
write.csv( water_analysis,  paste(gispath, collapse = ""), row.names = FALSE)
```

```{r second Housekeeping, include=FALSE}
rm(list=c("res.hc","hc1","water_analysis_clust","water_analysis_clust_scaled","map_data","myMap","numberofclusters","clusterCut","d"))
```

# 3. Data visualization

## 3.1. Piper plot

The code for the piper diagram is from [\<https://github.com/markolipka/ggplot_Piper\>](https://github.com/markolipka/ggplot_Piper)

```{r Functions for piper plot, include=FALSE}

library(ggplot2)

# this function will transform the data into 100%
transform_piper_data <- function(Mg, Ca, Cl,SO4, name=NULL){
  if(is.null(name)){
    name = rep(1:length(Mg),3)
  } else {
    name = rep(name,3)
  }
  y1 <- Mg * 0.86603
  x1 <- 100*(1-(Ca/100) - (Mg/200))
  y2 <- SO4 * 0.86603
  x2 <-120+(100*Cl/100 + 0.5 * 100*SO4/100)
  new_point <- function(x1, x2, y1, y2, grad=1.73206){
    b1 <- y1-(grad*x1)
    b2 <- y2-(-grad*x2)
    M <- matrix(c(grad, -grad, -1,-1), ncol=2)
    intercepts <- as.matrix(c(b1,b2))
    t_mat <- -solve(M) %*% intercepts
    data.frame(x=t_mat[1,1], y=t_mat[2,1])
  }
  np_list <- lapply(1:length(x1), function(i) new_point(x1[i], x2[i], y1[i], y2[i]))
  npoints <- do.call("rbind",np_list)
  data.frame(observation=name,x=c(x1, x2, npoints$x), y=c(y=y1, y2, npoints$y))
}

### This is the actual function to call the piper plot
 
ggplot_piper <- function() {
  library(ggplot2)
  grid1p1 <<- data.frame(x1 = c(20, 40, 60, 80),
                         x2 = c(10, 20, 30, 40),
                         y1 = c(0, 0, 0, 0),
                         y2 = c(17.3206, 34.6412, 51.9618, 69.2824)) ## FIXME: how are these numbers generated???
  grid1p2 <<- data.frame(x1 = c(20, 40, 60, 80),
                         x2 = c(60, 70, 80, 90),
                         y1 = c(0, 0, 0, 0),
                         y2 = c(69.2824, 51.9618, 34.6412, 17.3206)) ## FIXME: how are these numbers generated???
  grid1p3 <<- data.frame(x1 = c(10, 20, 30, 40), 
                         x2 = c(90, 80, 70, 60),
                         y1 = c(17.3206, 34.6412, 51.9618, 69.2824), ## FIXME: how are these numbers generated???
                         y2 = c(17.3206, 34.6412, 51.9618, 69.2824)) ## FIXME: how are these numbers generated???
  grid2p1 <<- grid1p1
  grid2p1$x1 <- grid2p1$x1 + 120
  grid2p1$x2 <- grid2p1$x2 + 120
  grid2p2 <<- grid1p2
  grid2p2$x1 <- grid2p2$x1 + 120
  grid2p2$x2 <- grid2p2$x2 + 120
  grid2p3 <<- grid1p3
  grid2p3$x1 <- grid2p3$x1 + 120
  grid2p3$x2 <- grid2p3$x2 + 120
  grid3p1 <<- data.frame(x1 = c(100, 90, 80, 70),
                         y1 = c(34.6412, 51.9618, 69.2824, 86.603), ## FIXME: how are these numbers generated???
                         x2 = c(150, 140, 130, 120),
                         y2 = c(121.2442, 138.5648, 155.8854, 173.2060)) ## FIXME: how are these numbers generated???
  grid3p2 <<- data.frame(x1 = c(70, 80, 90, 100),
                         y1 = c(121.2442, 138.5648, 155.8854, 173.2060), ## FIXME: how are these numbers generated???
                         x2 = c(120, 130, 140, 150),
                         y2 = c(34.6412, 51.9618, 69.2824, 86.603)) ## FIXME: how are these numbers generated???
  
  label.size <- 5
  
  p <- ggplot() +
    
    ## left hand ternary plot
    geom_segment(aes(x =  0, y =  0,     xend = 100, yend = 0)) +
    geom_segment(aes(x =  0, y =  0,     xend =  50, yend = 86.603)) + ## FIXME: how are these numbers generated???
    geom_segment(aes(x = 50, y = 86.603, xend = 100, yend = 0)) + ## FIXME: how are these numbers generated???
    
    ## right hand ternary plot
    geom_segment(aes(x = 120, y = 0, xend = 220, yend =  0)) +
    geom_segment(aes(x = 120, y = 0, xend = 170, yend = 86.603)) +
    geom_segment(aes(x = 170, y = 86.603, xend = 220, yend = 0)) +
    
    ## Upper diamond
    geom_segment(aes(x = 110, y = 190.5266, xend =  60, yend = 103.9236)) +
    geom_segment(aes(x = 110, y = 190.5266, xend = 160, yend = 103.9236)) +
    geom_segment(aes(x = 110, y =  17.3206, xend = 160, yend = 103.9236)) +
    geom_segment(aes(x = 110, y =  17.3206, xend =  60, yend = 103.9236)) +
    
    ## Add grid lines to the plots
    geom_segment(aes(x = x1, y = y1, yend = y2, xend = x2), data = grid1p1, linetype = "dashed", size = 0.25, colour = "grey50") +
    geom_segment(aes(x = x1, y = y1, yend = y2, xend = x2), data = grid1p2, linetype = "dashed", size = 0.25, colour = "grey50") +
    geom_segment(aes(x = x1, y = y1, yend = y2, xend = x2), data = grid1p3, linetype = "dashed", size = 0.25, colour = "grey50") +
    geom_segment(aes(x = x1, y = y1, yend = y2, xend = x2), data = grid2p1, linetype = "dashed", size = 0.25, colour = "grey50") +
    geom_segment(aes(x = x1, y = y1, yend = y2, xend = x2), data = grid2p2, linetype = "dashed", size = 0.25, colour = "grey50") +
    geom_segment(aes(x = x1, y = y1, yend = y2, xend = x2), data = grid2p3, linetype = "dashed", size = 0.25, colour = "grey50") +
    geom_segment(aes(x = x1, y = y1, yend = y2, xend = x2), data = grid3p1, linetype = "dashed", size = 0.25, colour = "grey50") +
    geom_segment(aes(x = x1, y = y1, yend = y2, xend = x2), data = grid3p2, linetype = "dashed", size = 0.25, colour = "grey50") +
    
    ### Labels and grid values
    coord_fixed(ratio = 1) +  
    geom_text(aes(17,   50, label = "Mg^'2+'"), angle = 60, size = label.size, parse = TRUE) +  
    geom_text(aes(82.5, 50, label = "Na^'+'~+~K^'+'"), angle = -60, size = label.size, parse = T) +
    geom_text(aes(50,  -10, label = "Ca^'2+'"), size = label.size, parse = TRUE) +
    geom_text(aes(170,   -10, label = "Cl^'-'"), size = label.size, parse = TRUE) +
    geom_text(aes(205,    50, label = "SO[4]^'2-'"), angle = -60, size = label.size, parse = TRUE) +
    geom_text(aes(137.5,  50, label = "Alkalinity~as~HCO[3]^'-'"), angle = 60, size = label.size, parse = TRUE) +
    geom_text(aes( 72.5, 150, label = "SO[4]^'2-'~+~Cl^'-'"), angle = 60, size = label.size, parse = TRUE) +
    geom_text(aes(147.5, 150, label = "Ca^'2+'~+~Mg^'2+'"), angle = -60, size = label.size, parse = TRUE) + 
    
    geom_text(aes(c(35, 25, 15, 5), grid1p2$y2, label = c(80, 60, 40, 20)), size = label.size -1, angle = 0) + # Mg axis
    geom_text(aes(c(95, 85, 75, 65), grid1p3$y2, label = c(80, 60, 40, 20)), size = label.size -1, angle = 60, vjust = -1, hjust = 0) + # Na axis
    geom_text(aes(c(20, 40, 60, 80), c(-5, -5, -5, -5), label = c(80, 60, 40, 20)), size = label.size -1, angle = -60, vjust = -.5) + # Ca axis
    geom_text(aes(c(155, 145, 135, 125), grid2p2$y2, label = c(20, 40, 60, 80)), size = label.size -1, angle = -60, vjust = -1, hjust = 1) + # HCO3 axis
    geom_text(aes(c(215, 205, 195, 185), grid2p3$y2, label = c(20, 40, 60, 80)), size = label.size -1, angle = 0) + # SO4 axis
    geom_text(aes(c(140, 160, 180, 200), c(-5, -5, -5, -5), label = c(20, 40, 60, 80)), size = label.size -1, angle = 60, vjust = -.5) + # Cl axis
    #geom_text(aes(grid3p1$x1 - 5, grid3p1$y1, label = c(80, 60, 40, 20)), size=3, angle = 60, vjust = -1.5, hjust = 1) + # diamond Na axis
    geom_text(aes(grid3p1$x2 + 5, grid3p1$y2, label = c(20, 40, 60, 80)), size = label.size -1, angle =  60, vjust = -1, hjust = 0) + # diamond Ca axis
    geom_text(aes(grid3p2$x1 - 5, grid3p2$y1, label = c(20, 40, 60, 80)), size = label.size -1, angle = -60, vjust = -1, hjust = 1) + # diamond SO4 axis
    #geom_text(aes(grid3p2$x2 + 5, grid3p2$y2, label = c(80, 60, 40, 20)), size=3, angle =  90) + # diamond HCO3 axis
    theme_bw() +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
          panel.border = element_blank(), axis.ticks = element_blank(),
          axis.text.x = element_blank(), axis.text.y = element_blank(),
          axis.title.x = element_blank(), axis.title.y = element_blank())
  return(p)
}
```

```{r Data preperation and piper plot, echo=FALSE, fig.height=10, fig.width=10, fig.cap="Piper plot of the hydrochemical data from Samos"}
library(ggplot2)
library(ggrepel)
library("hydrogeo")

if(exists("CO3_meq")){
   CO3 = CO3_meq 
 }else{
   CO3 = replicate(length(water_analysis$Sample), 0)
 }

milliequivalents = list( Ca   = Ca_meq,
                         Mg   = Mg_meq,
                         Na   = Na_meq,
                         K    = K_meq,
                         Cl   = Cl_meq,
                         SO4  = SO4_meq,
                         CO3 = CO3,
                         HCO3 = HCO3_meq,
                         "Samples" = water_analysis$Sample,
                         "TDS" = water_analysis$TDS,
                         "Cluster" = water_analysis$Cluster)

### This transforms the meq/L to percents 
percents <- toPercent(milliequivalents)

data <- as.data.frame(percents)

### This applies plot coordinates to the variables
piper_data <- transform_piper_data(Ca   = data$Ca,
                                   Mg   = data$Mg,
                                   Cl   = data$Cl,
                                   SO4  = data$SO4,
                                   name = data$Samples)

piper_data <- merge(piper_data,
                     data[, c("TDS", "Cluster", "Samples")],
                     by.y = "Samples",
                     by.x = "observation")

#######

piper_data$Cluster= as.factor(piper_data$Cluster)
piper_labels = as.character(piper_data$observation)
piper_data$label = piper_labels
rm(piper_labels)

#### Samples, TDS, Clusters

piper_2 = ggplot_piper() + geom_point(aes(x,y,colour=factor(Cluster), size = TDS), data=piper_data) + 
  labs(color = "Cluster") + 
  theme(legend.position="bottom")+
  geom_text_repel(data =piper_data, aes(x, y, label = label), size = 3, max.overlaps = 30, alpha = 0.5 )

piper_2 + scale_size_area()+ scale_size_continuous(range = c(1, 10))

ggsave(file="Piper.svg", piper_2, width=10, height=10)

```

## 3.2. Shoeller plot

```{r Reorder milliequivalents data frame for the shoeller plot, include=FALSE,}

library(tidyverse)

milliequivalents = as.data.frame(milliequivalents)
milliequivalents$Cluster = as.factor(milliequivalents$Cluster)

if (exists("NO3_meq")) {
  milliequivalents$NO3 = NO3_meq
}

if (exists("Br_meq")) {
  milliequivalents$Br =  Br_meq
}

if (exists("CO3_meq")) {
  milliequivalents$CO3 =  CO3_meq
}

### Elements and their order to be shown in shoeller plot.
milliequivalents_shoeller= milliequivalents%>% 
  select(Samples,Cluster,TDS,K,Na,Cl,Ca,SO4,Mg,HCO3)

### Copy the Ions (not Samples, Cluster, TDS) and place assign them as character (e.g. "Na", "cl" ) to the idmeasure.vars variable as instructed in the next chunk! 
```

Plot the Shoeller plot

```{r Schoeller Diagram, echo=FALSE, fig.height=12, fig.width=10, warning=FALSE, fig.cap=" Shoeller plot of the hydrochemical data from Samos" }

library(dplyr)
library(directlabels)
library(reshape2)
library(scales)
library(ggplot2)
library(plotly)

shoeller_data = subset(milliequivalents_shoeller, select=-c(TDS))
###
### HERE : assign the ions you want to plot to the idmeasure.vars variable
###
melted_shoeller_data = melt(shoeller_data,id.vars=c("Cluster","Samples"),idmeasure.vars = c("K","Na","Cl","Ca","SO4","Mg","HCO3"))

###  
shoeller_plot= ggplot(data = melted_shoeller_data)+
  geom_line(aes(x = variable, y = value, group = Samples, colour = Cluster), alpha = 0.8)+
  geom_dl(aes(x = variable, y = value,label = Samples),method = list(dl.combine("first.qp","last.qp")), cex = 2, size= 0.01)+
  geom_point(aes(x = variable, y = value,colour = Cluster, group = Samples), alpha= 0.5)+
  scale_x_discrete()+
  theme_light()+
  scale_y_continuous(trans = scales::log10_trans(),
                     breaks = scales::pretty_breaks(),
                     labels = prettyNum)+
  labs(y = "meq/L", x= "")+
  theme(legend.position="bottom")+
  annotation_logticks(sides = "lr")

ggsave(file= "Shoeller_all.svg", shoeller_plot , width=10, height=12)

shoeller_plot  
```

The mean value for every variable in each cluster is calculated and plotted again in a Shoeller diagram.

```{r Mean value for each cluster and mean shoeller plot, echo=FALSE, fig.cap="Shoeller plot presenting the *average* concentration of each cluster", fig.height=7, message=FALSE, warning=FALSE}

library(knitr)
library(ggplot2)
### Choose the numeric elements to be averaged for each cluster
summary_milliequivalents = subset(shoeller_data, select=-c(Samples))

mean_values_of_clusters = aggregate(. ~ Cluster, data = summary_milliequivalents, FUN = function(x) c( mn = mean(x) ) )

kable(mean_values_of_clusters,align = 'c', caption = 'Mean values for each cluster ($meq/L$)')



mean_values_of_clusters = melt(mean_values_of_clusters,id.vars=c("Cluster"),idmeasure.vars = c("K","Na","Cl","Ca","SO4","Mg","HCO3"))

shoeller_avg = ggplot(data = mean_values_of_clusters)+
  geom_line(aes(x = variable, y = value, group = Cluster, colour = Cluster), alpha = 0.8)+
  geom_point(aes(x = variable, y = value,colour = Cluster, group =Cluster))+
  scale_x_discrete()+
  theme_light()+
  scale_y_continuous(trans = scales::log10_trans(),
                     breaks = scales::pretty_breaks(),
                     labels = prettyNum)+
  labs(y = "meq/L", x= "")+
  theme(legend.position="bottom")+
  annotation_logticks(sides = "lr")

ggsave(file="Shoeller_cluster_average.svg", shoeller_avg , width=8, height=5)

shoeller_avg
```

The variables plotted in the Shoeller plots can be summarized in boxplots.

```{r Summarizing boxplots, echo=FALSE, fig.height=6, fig.cap= " Boxplots of the clusters"}

library(ggplot2)
library(scales)

### same y axis
box_com=ggplot(data = melted_shoeller_data, mapping = aes(x = variable, y = value, color= Cluster)) + 
  geom_boxplot()+
  scale_x_discrete()+
  theme_light()+
  labs(y = "Common scale: in meq/L", x= "")+
  theme(legend.position="bottom")+
  facet_wrap(~Cluster)

ggsave(file="Boxplot_common_scale.svg", box_com , width=8, height=5)
box_com
### free y axis
free_com =ggplot(data = melted_shoeller_data, mapping = aes(x = variable, y = value, color= Cluster)) + 
  geom_boxplot()+
  scale_x_discrete()+
  theme_light()+
  labs(y = "Free scale: meq/L", x= "")+
  theme(legend.position="bottom")+
  facet_wrap(~Cluster, scales = "free")

ggsave(file="Boxplot_free_scale.svg", free_com , width=8, height=5)
free_com
```

## 3.4. Composite diagrams

```{r Na-Cl, echo=FALSE, fig.cap=""}
library(ggplot2)
library(scales)
library(latex2exp)
library(ggforce)

Na_to_Cl_plot =ggplot(data=milliequivalents)+
  geom_point(aes(x = Cl , y = Na, color = Cluster))+
  geom_abline(intercept = 0, slope = 1, color = "grey", linetype = "dashed")+
  theme_light()+
  xlab(TeX(r'($Cl ^{-}\;meq/L$)')) +
  ylab(TeX(r'($Na ^{+}\;meq/L$)')) +
  
  
  ### adds zoomed window 
  facet_zoom(ylim = c(0, 30), xlim = c(0, 40))


ggsave(file="Na_vs_Cl.svg", Na_to_Cl_plot)
Na_to_Cl_plot
```

```{r Br-Cl, echo=FALSE, fig.cap=""}
library(ggplot2)
library(scales)
library(latex2exp)

if(exists("Br_meq")) {
  Br_to_Cl=  ggplot(data = milliequivalents) +
    geom_point(data = subset(milliequivalents, Br > 0),aes(x = Cl , y = Br, color = Cluster)) +
    geom_abline(
      intercept = 0,
      slope = 1,
      color = "grey",
      linetype = "dashed"
    ) +
    theme_light() +
   xlab(TeX(r'($Cl ^{-}\;meq/L$)')) +
  ylab(TeX(r'($Br ^{-}\;meq/L$)'))+
  # Delete this part if you don't want log axes
    scale_y_continuous(trans = scales::log10_trans(),
                   breaks = scales::pretty_breaks(),
                   labels = prettyNum)+
  scale_x_continuous(trans = scales::log10_trans(),
                     breaks = scales::pretty_breaks(),
                     labels = prettyNum)
}

if(exists("Br_vs_Cl")){ggsave(file="Br_vs_Cl.svg",  Br_to_Cl)}
```

```{r Ca to SO4, echo=FALSE, fig.cap=""}
library(ggplot2)
library(scales)
library(latex2exp)


Ca_to_SO4_plot =ggplot(data=milliequivalents)+
  geom_point(aes(x = SO4 , y = Ca, color = Cluster))+
  geom_abline(intercept = 0, slope = 1, color = "grey", linetype = "dashed")+
  theme_light()+
   xlab(TeX(r'($SO4 ^{2-}\;meq/L$)')) +
  ylab(TeX(r'($ Ca^{2+}\;meq/L$)'))+
   facet_zoom(ylim = c(0,15), xlim = c(0,15))
  

ggsave(file="Ca_vs_SO4.svg",  Ca_to_SO4_plot)
Ca_to_SO4_plot 
```

```{r Na_plus_K_to_NaMgCaK, echo=FALSE, fig.cap=""}

library(ggforce)

Na_plus_K_to_NaMgCaK =ggplot(data=milliequivalents)+
  geom_point(aes(x = (Na+Mg+K+Ca) , y = Na+K, color = Cluster))+
  geom_abline(intercept = 0, slope = c(1,0.5,1.5), color = "grey", linetype = "dashed")+
  theme_light()+
  labs(y = "[Na + K] meq/L", x= "[Na+K+Mg+Ca] meq/L") +

  ### adds zoomed window 
  facet_zoom(ylim = c(0, 50), xlim = c(0,50))




ggsave(file="Na_plus_K_vs__NaMgCaK.svg",  Na_plus_K_to_NaMgCaK)
Na_plus_K_to_NaMgCaK
```

```{r Ca_plus_Mg_to_HCO3, echo=FALSE, fig.cap=""}
library(ggplot2)
library(latex2exp)
library(ggforce)

Ca_plus_Mg_to_HCO3 =ggplot(data=milliequivalents)+
  geom_point(aes(x = HCO3 , y = (Ca+Mg), color = Cluster))+
  geom_abline(intercept = 0, slope = 1, color = "grey", linetype = "dashed")+
  theme_light()+
   xlab(TeX(r'($HCO3^{-}\;meq/L$)')) +
  ylab(TeX(r'($ Ca + Mg\;meq/L$)')) +
  
 
  ### adds zoomed window 
  facet_zoom(ylim = c(0, 30), xlim = c(0, 10))

ggsave(file="Ca_plus_Mg_vs_HCO3.svg",  Ca_plus_Mg_to_HCO3)
Ca_plus_Mg_to_HCO3

```

```{r Cl_to_HCO3, echo=FALSE, fig.cap=""}
library(ggplot2)
library(latex2exp)
library(ggforce)

Cl_to_HCO3 =ggplot(data=milliequivalents)+
  geom_point(aes(x =HCO3 , y = Cl, color = Cluster))+
  geom_abline(intercept = 0, slope = 1, color = "grey", linetype = "dashed")+
  theme_light()+
  labs(y = "Cl meq/L", x= "HCO3 meq/L")+
  xlab(TeX(r'($HCO3^{-}\;meq/L$)')) +
  ylab(TeX(r'($Cl ^{-}\;meq/L$)'))+
  
  facet_zoom(ylim = c(0, 30), xlim = c(0, 7))
  

ggsave(file="Cl_vs_HCO3.svg",  Cl_to_HCO3)

Cl_to_HCO3
```

```{r Arrange composite plots into a grid, eval=FALSE, include=FALSE, fig.cap=}
library(ggpubr)

composite_plot_grid = ggarrange(
  Na_to_Cl_plot, Ca_to_SO4_plot, Na_plus_K_to_NaMgCaK, Ca_plus_Mg_to_HCO3, Cl_to_HCO3,
  ncol = 3, 
  nrow = 2, labels = c("A", "B", "C", "D", "E"),
  common.legend = TRUE, legend = "bottom", heights = 10, font.label = list(size = 10)) 

ggsave(file="Composite_chemical_ratios.svg",  composite_plot_grid)

composite_plot_grid
```

## 3.5. Isotopic data

```{r 2H to 18O, echo=FALSE, fig.cap=""}
library(ggplot2)
library(latex2exp)
library(scales)

if(("d18O" %in% colnames(water_analysis)) && ("d2H" %in% colnames(water_analysis)) ) {

GMWL <- function(x) {
  8 * x + 10
}
MMWL <- function(x) {
  8 * x + 20
}


deuterium_to_18O = ggplot(data = water_analysis)  + stat_function(
  data = data.frame(x = 0),
  aes(x = x, col = "GMWL", linetype = "GMWL"),
  col = "black",
  fun = GMWL
) +
  stat_function(
    data = data.frame(x = 0),
    aes(x = x, col = "MMWL", linetype = "MMWL"),
    col = "black",
    fun = MMWL
  ) +
  geom_point(aes(
    x = d18O
,
    y = d2H
,
    stroke = 0.1,
    color = as.factor(Cluster)
  ), alpha = 0.9) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  xlab(expression(paste(delta ^ {
    18
  }, "O(‰, V-SMOW)"))) +
  ylab(expression(paste(delta ^ {
    2
  }, "H(‰, V-SMOW)"))) +
  theme_light() +
  theme(legend.position = "right", panel.grid.minor = element_blank()) +
  labs(colour = "Cluster",
       linetype = "")

ggsave(file = "d18O_2H.svg",deuterium_to_18O,width = 10,height = 8)}

if(exists("deuterium_to_18O")){
  deuterium_to_18O
}


```

```{r DE to 18O, echo=FALSE, fig.cap=""}
library(ggplot2)
library(scales)

 if(("d18O" %in% colnames(water_analysis)) && ("D_Excess" %in% colnames(water_analysis)) ) {

de_to_O18 = ggplot(data = water_analysis) +
  geom_point(aes(
    x = d18O
,
    y = D_Excess
,
    stroke = 0.1,
    color = as.factor(Cluster)
  ), alpha = 0.9) +
  xlab(expression(paste(delta ^ {
    18
  }, "O(‰, V-SMOW)"))) +
  ylab(expression(paste("DE(‰)"))) +
  theme_light() +
  theme(legend.position = "right", panel.grid.minor = element_blank()) +
  labs(colour = "Cluster",
       linetype = "")

ggsave(file = "DE_18O.svg",de_to_O18,width = 10,height = 8)}


if(exists("de_to_O18")){
  de_to_O18
}

```

```{r cl_to_O18, echo=FALSE, fig.cap=""}
library(latex2exp)
library(ggplot2)
library(scales)

if("d18O" %in% colnames(water_analysis)) {
o18_to_cl = ggplot(data = water_analysis) +
  geom_point(aes(
    x = log(milliequivalents$Cl),y =d18O,stroke = 0.1,color = as.factor(Cluster)), alpha = 0.6) +
  ylab(expression(paste(delta ^ {18}, "O(‰, V-SMOW)"))) +
  xlab(TeX(r'($log(Cl ^{-})\;meq/L$)'))+ 
  theme_light() +
  theme(legend.position = "right", panel.grid.minor = element_blank()) +
  labs(colour = "Cluster",
           linetype = "")

ggsave(file = "Cl_18O.svg", o18_to_cl ,width = 10,height = 8)
}

if(exists("o18_to_cl")){
 
  o18_to_cl 
  
  }
```

```{r DE_to_Cl, echo=FALSE, fig.cap=""}
library(latex2exp)
library(ggplot2)
library(scales)

if("D_Excess" %in% colnames(water_analysis)) {
de_to_cl = ggplot(data = water_analysis) +
  geom_point(aes(
    x = log(milliequivalents$Cl),y = D_Excess,stroke = 0.1,color = as.factor(Cluster)), alpha = 0.6) +
  ylab(expression(paste("DE(‰)"))) +
  xlab(TeX(r'($log(Cl ^{-})\;meq/L$)'))+ 
  theme_light() +
  theme(legend.position = "right", panel.grid.minor = element_blank()) +
  labs(colour = "Cluster",
           linetype = "")

ggsave(file = "DE_Cl.svg", de_to_cl,width = 10,height = 8) 
}

if(exists("de_to_cl")){
 
  de_to_cl
  
  }

```
